---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 6 - Due date 03/26/21"
author: "Rajat Khandelwal"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp21.Rmd"). Submit this pdf using Sakai.

## Set up

```{r warnings = FALSE, messages = FALSE, errors = FALSE, info = FALSE}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(readxl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyverse)
library(readr)
```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r}
setwd("/Users/rajatkhandelwal/Documents/GitHub/ENV790_30_TSA_S2021")
data <- read_csv("Data/Net_generation_United_States_all_sectors_monthly.csv", skip = 4) 
data <- data %>% rename ("Natural Gas Generation (in 1000s of MWh)" = "natural gas thousand megawatthours") %>% select(1,4)
ncols = ncol(data) - 1
nobs = nrow(data)
head(data)
data$Month <- my(data$Month)
str(data)
data_ts <- ts(data[2], end = c(year(data$Month[1]), month(data$Month[1])), start = c(year(data$Month[nobs]),month(data$Month[nobs])),frequency = 12)

ggplot(data, aes(data$Month,data_ts)) +
  geom_line(color = "red")
  
par(mfrow = c(1,2))
Acf(data_ts)
Pacf(data_ts)

```

### Q2

Using the $decompose()$ or $stl()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the deseasonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.

```{r}
data_ts_decomp <- decompose(data_ts)
data_ts_deseasoned <- seasadj(data_ts_decomp)

ggplot(data, aes(data$Month,data_ts_deseasoned)) +
  geom_line(color = "red")
  
par(mfrow = c(1,2))
Acf(data_ts_deseasoned)
Pacf(data_ts_deseasoned)
```

## Modeling the seasonally adjusted or deseasonalized series

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

```{r}
print("ADF Test")
print(adf.test(data_ts_deseasoned))
print("Mann-Kendall test")
print(MannKendall(data_ts_deseasoned))
```
Since the p-value for for the ADF test is < 0.05, we reject the null hypothesis that the time-series has a unit root. Hence, we conclude that the time-series is stationary. Further, since the p-value of the Mann-Kendall test is < 0.05 hence we again reject the null hypothesis that there is no deterministic trend in the series.

Therefore, based on these results, we can state that the deseasonal time-series does not have a stochastic trend but it does have a deterministic trend.


### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to can read the plots and interpret the test results.

```{r}
ndiff <- ndiffs(data_ts_deseasoned)
ndiff
```
p = 2 (barely)
q = 0
d = 1 since we have a determinsitic trend

### Q5

Use $Arima()$ from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. Should you allow for constants in the model, i.e., $include.mean = TRUE$ or $include.drift=TRUE$. **Print the coefficients** in your report. Hint: use the $cat()$ function to print.

```{r}
data_ts_deseasoned_arima <- Arima(data_ts_deseasoned, order = c(2,1,0), seasonal = c(0,0,0), include.mean = FALSE, include.constant = FALSE)

```
cat("The value of $\phi_{1}$ is %.1f", data_ts_deseasoned_arima[[1]][1])

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF and PACF on the same window. You may use the $checkresiduals()$ function to automatically generate the three plots. Do the residual series look like a white noise series? Why?

## Modeling the original series (with seasonality)

### Q7

Repeat Q4-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.

```{r}

```


### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.

```{r}

```


## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the correct orders. The intention of the assignment is to walk you to the process and help you figure out what you did wrong (if you did anything wrong!).

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}

```


### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}

```

